Gravitational-wave sensitivity curves

### 2014-09-26 V2/ReplyToReferees.txt

Other comments (in order of appearance in the text):
1)	We thank the referee for pointing out this interesting source of confusion. We have checked this, and it seems that it was originally "evolved LISA", but the alternative "evolving LISA" has gradually become more popular and is now the prefered option. We have found numerous examples of both acronyms being used; for two conflicting examples on the same website see https://www.elisascience.org/news/top-news/gravitationaluniverseselectedasl3 and https://www.elisascience.org/articles/elisa-consortium/home . We have changed the main body of the text to read "evolving" added the footnote:

"The e in eLISA originally stood for evolved, but the preference now is for evolving."

2)	We agree with the referees comment and have amended the passage to read:

"Another disadvantage is that applying the definition 17 to a monochromatic source gives a formally undefined answer. The correct identification of characteristic strain for a monochromatic source is the amplitude of the wave times the square root of the number of periods observed (see section \ref{sec:insp})."

### 2014-09-26 V2/Detectors.tex
final tweaks

The next great advancement in radio astronomy shall come with the completion of the Square Kilometre Array \citep[SKA;][]{Dewdney2009}. This shall greatly increase the sensitivity of pulsar timing \citep{Kramer2004}. The sensitivity curves plotted in \ref{app:a} for the SKA assumes $50$ pulsars timed every 2 weeks for 15 years with an r.m.s.\ error in each timing residual of $100\,\textrm{ns}$. Our choice of 50 pulsars might be conservative, it is possible that the SKA will discover many more suitable milli-second pulsars than this; however, not all pulsars will be discovered soon enough to accumulate sufficient observations to be of use in an analysis. %% CPLB: rewording, please check!

>>>>

The next great advancement in radio astronomy shall come with the completion of the Square Kilometre Array \citep[SKA;][]{Dewdney2009}. This shall greatly increase the sensitivity of pulsar timing \citep{Kramer2004}. The sensitivity curves plotted in \ref{app:a} for the SKA assumes $50$ pulsars timed every 2 weeks for 15 years with an r.m.s.\ error in each timing residual of $100\,\textrm{ns}$. Our choice of 50 pulsars might be conservative, it is possible that the SKA will discover many more suitable milli-second pulsars than this during its operation, in this case the corresponding sensitivity curve would be lower.

### 2014-09-26 V2/Introduction.tex
final tweaks

Given a detector output, the challenge is to extract the signal. There is a well known solution to this problem that involves constructing a Wiener optimal filter \citep{Wiener49}. Let $K(t)$ be a real filter function with Fourier transform $\tilde{K}(f)$. Convolving this with the detector output gives a contribution from the signal and a contribution from the noise,
\begin{equation}\label{eq:conv}
\left(s*K\right)(\tau) = \int_{-\infty}^{\infty}\mathrm{d}t\;\left[h(t)+n(t)\right]K(t-\tau) \approx \mathcal{S} + \mathcal{N} \; .
\end{equation}
The signal contribution $\mathcal{S}$ is defined as the expectation of the convolution in (\ref{eq:conv}) when a signal is present, maximised by varying the offset to achieve the best overlap with the data. Since the expectation of pure noise is zero it follows that
\begin{equation}
\mathcal{S} = \int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K(t)=\int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K^{*}(t)=\int_{-\infty}^{\infty}\mathrm{d}f\; \tilde{h}(f)\tilde{K}^{*}(f) \; .
\end{equation}
The squared contribution from noise $\mathcal{N}^{2}$ is defined as the mean square of the convolution in (\ref{eq:conv}) when no signal is present,
\begin{eqnarray} 
\mathcal{N}^{2} &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K(t')\left<n(t)n(t')\right> \nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K^{*}(t')\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\exp\left[2\pi\rmi(ft-f't')\right]\nonumber \\
%&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;\frac{1}{2}S_{n}(f)K(t)K^{*}(t')\exp\left(2\pi\rmi f (t-t')\right)\nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}f \; \frac{1}{2}S_{n}(f)\tilde{K}(f)\tilde{K}^{*}(f) \; ,
 \end{eqnarray}
using the definition of $S_{n}(f)$ from (\ref{eq:psd}). Hence the signal-to-noise ratio (SNR) $\varrho$ is given by
\begin{equation}\label{eq:SNRinnerprod} 
\varrho^{2} = \frac{\mathcal{S}^{2}}{\mathcal{N}^{2}}= \frac{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\tilde{h}(f)}^{2}}{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\frac{1}{2}S_{n}(f)\tilde{K}(f)}},
\end{equation}
where we have introduced the inner product between signal $\tilde{A}$ and $\tilde{B}$ as \citep{Finn1992}
\begin{equation}\label{eq:snrinnerprod} \innerprod{\tilde{A}(f)}{\tilde{B}(f)} = 4\Re\left\{\int_{0}^{\infty}\mathrm{d}f\;\frac{\tilde{A}^{*}(f)\tilde{B}(f)}{S_{n}(f)}\right\} \; .\end{equation}
The optimum filter is that function $\tilde{K}(f)$ which maximises the SNR in (\ref{eq:SNRinnerprod}). From the Cauchy--Schwarz inequality, it follows that the optimum filter is
\begin{equation}
\tilde{K}(f)=\frac{\tilde{h}(f)}{S_{n}(f)} \; .
\end{equation}
This is the Wiener filter, which may be multiplied by an arbitrary constant since this does not change the SNR. Using this form for $\tilde{K}(f)$, the squared SNR is
\begin{equation}
\varrho^2 = \int_0^\infty\mathrm{d}f \frac{4| \tilde{h}(f)|^{2}}{S_n(f)} = \innerprod{\tilde{h}(f)}{\tilde{h}(f)}.
\label{eq:traditionalSNR} 
\end{equation}
In order to construct the Wiener filter, it is necessary to know \emph{a priori} the form of the signal, $\tilde{h}(f)$, for this reason the Wiener filter is sometimes called the \emph{matched} filter.

@@

An additional advantage of this convention is that the values on the strain axis for the detector curve $h_n(f)$ have a simple physical interpretation: they correspond to the root-mean-square noise in a bandwidth $f$. One downside to plotting characteristic strain is that the values on the strain axis $h_\mathrm{c}(f)$ do not directly relate to the amplitude of the waves from the source. 
Another disadvantage is that applying (\ref{eq:strain-hc}) to a monochromatic source gives a formally undefined answer. The correct identification of characteristic strain for a monochromatic source is the amplitude of the wave times the square root of the number of periods observed. %% CPLB: Didn't we used to have a section explaining this? Perhaps add ref to 3.1?

>>>>

Given a detector output, the challenge is to extract the signal. There is a well known solution to this problem that involves constructing a Wiener optimal filter \citep{Wiener49}. Let $K(t)$ be a real filter function with Fourier transform $\tilde{K}(f)$. Convolving this with the detector output gives a contribution from the signal and a contribution from the noise,
\begin{equation}\label{eq:conv}
\left(s*K\right)(\tau) = \int_{-\infty}^{\infty}\mathrm{d}t\;\left[h(t)+n(t)\right]K(t-\tau) \approx \mathcal{S} + \mathcal{N} \; .
\end{equation}
The signal contribution $\mathcal{S}$ is defined as the expectation of the convolution in (\ref{eq:conv}) when a signal is present, maximised by varying the offset to achieve the best overlap with the data. Since the expectation of pure noise is zero it follows that
\begin{equation}
\mathcal{S} = \int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K(t)=\int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K^{*}(t)=\int_{-\infty}^{\infty}\mathrm{d}f\; \tilde{h}(f)\tilde{K}^{*}(f) \; .
\end{equation}
The squared contribution from noise $\mathcal{N}^{2}$ is defined as the mean square of the convolution in (\ref{eq:conv}) when no signal is present,
\begin{eqnarray} 
\mathcal{N}^{2} &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K(t')\left<n(t)n(t')\right> \nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K^{*}(t')\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\exp\left[2\pi\rmi(ft-f't')\right]\nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}f \; \frac{1}{2}S_{n}(f)\tilde{K}(f)\tilde{K}^{*}(f) \; ,
 \end{eqnarray}
using the definition of $S_{n}(f)$ from (\ref{eq:psd}). Hence the signal-to-noise ratio (SNR) $\varrho$ is given by
\begin{equation}\label{eq:SNRinnerprod} 
\varrho^{2} = \frac{\mathcal{S}^{2}}{\mathcal{N}^{2}}= \frac{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\tilde{h}(f)}^{2}}{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\frac{1}{2}S_{n}(f)\tilde{K}(f)}},
\end{equation}
where we have introduced the inner product between signal $\tilde{A}$ and $\tilde{B}$ as \citep{Finn1992}
\begin{equation}\label{eq:snrinnerprod} \innerprod{\tilde{A}(f)}{\tilde{B}(f)} = 4\Re\left\{\int_{0}^{\infty}\mathrm{d}f\;\frac{\tilde{A}^{*}(f)\tilde{B}(f)}{S_{n}(f)}\right\} \; .\end{equation}
The optimum filter is that function $\tilde{K}(f)$ which maximises the SNR in (\ref{eq:SNRinnerprod}). From the Cauchy--Schwarz inequality, it follows that the optimum filter is
\begin{equation}
\tilde{K}(f)=\frac{\tilde{h}(f)}{S_{n}(f)} \; .
\end{equation}
This is the Wiener filter, which may be multiplied by an arbitrary constant since this does not change the SNR. Using this form for $\tilde{K}(f)$, the squared SNR is
\begin{equation}
\varrho^2 = \int_0^\infty\mathrm{d}f \frac{4| \tilde{h}(f)|^{2}}{S_n(f)} = \innerprod{\tilde{h}(f)}{\tilde{h}(f)}.
\label{eq:traditionalSNR} 
\end{equation}
In order to construct the Wiener filter, it is necessary to know \emph{a priori} the form of the signal, $\tilde{h}(f)$, for this reason the Wiener filter is sometimes called the \emph{matched} filter.

@@

An additional advantage of this convention is that the values on the strain axis for the detector curve $h_n(f)$ have a simple physical interpretation: they correspond to the root-mean-square noise in a bandwidth $f$. One downside to plotting characteristic strain is that the values on the strain axis $h_\mathrm{c}(f)$ do not directly relate to the amplitude of the waves from the source. 
Another disadvantage is that applying (\ref{eq:strain-hc}) to a monochromatic source gives a formally undefined answer. The correct identification of characteristic strain for a monochromatic source is the amplitude of the wave times the square root of the number of periods observed (see section \ref{sec:insp}).
